A weakly structured stem for human origins in Africa

### 2022-07-15 paper.tex
small edits to address reviewer comments

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
479ka. The timing of this divergence was also poorly constrained ($\sigma_T=
166$ka). These populations evolve independently until approximately 119ka
when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the western Africans and
eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25ka.
The later Stem 2 contribution to the western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
This may indicate that an ancestral Stem 2 population occupied western or
Central Africa, broadly speaking. The differing proportions in the Nama and
eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred models. Many 
previous studies have found a reduction of coalescence rates 
between $1000$ka and $100$ka in humans, and thus inferred an increase 
in $N_e$ during the same period \citep{Li2011-le}.
This increase in inferred $N_e$ could be due to either an increase in population size or to 
ancestral population structure during the Middle Stone Age \citep{Mazet2015-wn}.
All models, including the
single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between 100ka-1Ma (Figure~\ref{fig:supp-iicr-sim}).
Whereas the single-origin model achieves this by
an increase in $N_e$ during that period, the best-fit model recapitulate
this pattern without corresponding population size changes.      
%\comment{These are pretty complicated ideas that are not explained in detail.
%Might be worth unpacking a bit more to make sure the reader understands}

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single origin model must be explained by archaic
hominin admixture. This is also true of this study. While it remains prohibitive to
fully explore the space of plausible models of early human population structure,
we sought to capture model uncertainty by exploring multiple parameterizations of
early history.
The best-fit models presented here include reticulation and migration between
early human populations rather than archaic hominin admixture from long-isolated branches.
We cannot rule out that more complex models involving
additional stems, or hybrid models including both weak structure and archaic
hominin admixture may better explain the data. Because parameters related to the split
time, migration rates, and relative sizes of the early stems were variable
across models, reflecting a degree of confounding among these parameters, we
refrained from introducing additional branches associated with more
parameters during that period.
Rather than interpreting the two stems as representing well-defined
and stable populations over hundreds of thousands of years, we interpret the
weakly structured stem as consistent with a population coalescence and
fragmentation model \citep{Scerri2019-xg}.
Models including additional diversity within Africa,
and early ancient DNA samples from Africa, could further distinguish
the archaic hominin admixture model from the weakly-structured-stem model.

>>>>

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
479ka. The timing of this divergence was also poorly constrained ($\sigma_T=
166$ka). These populations evolve independently until approximately 119ka
when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the western Africans and
eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25ka.
The later Stem 2 contribution to the western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
Central Africa, broadly speaking. The differing proportions in the Nama and
eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred models. Many 
previous studies have found a reduction of coalescence rates 
between $1000$ka and $100$ka in humans, and thus inferred an increase 
in $N_e$ during the same period \citep{Li2011-le}.
This increase in inferred $N_e$ could be due to either an increase in population size or to 
ancestral population structure during the Middle Stone Age \citep{Mazet2015-wn}.
All models, including the
single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between 100ka-1Ma extending deeper in the past than the relate estimate (Figure~\ref{fig:supp-iicr-sim} and Section \ref{inferred_relate}).
Whereas the single-origin model achieves this by
an increase in $N_e$ during that period, the best-fit model recapitulate
this pattern without corresponding population size changes.

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single origin model must be explained by archaic
hominin admixture. This is also true of this study. While it remains prohibitive to
fully explore the space of plausible models of early human population structure,
we sought to capture model uncertainty by exploring multiple parameterizations of
early history.
The best-fit models presented here include reticulation and migration between
early human populations rather than archaic hominin admixture from long-isolated branches.
We cannot rule out that more complex models involving
additional stems, continuous structure, or hybrid models including both weak structure and archaic
hominin admixture may better explain the data. Because parameters related to the split
time, migration rates, and relative sizes of the early stems were variable
across models, reflecting a degree of confounding among these parameters, we
refrained from introducing additional branches associated with more
parameters during that period.
Rather than interpreting the two stems as representing well-defined
and stable populations over hundreds of thousands of years, we interpret the
weakly structured stem as consistent with a population coalescence and
fragmentation model \citep{Scerri2019-xg}.
Models including additional diversity within Africa,
and early ancient DNA samples from Africa, could further distinguish
the archaic hominin admixture model from the weakly-structured-stem model.
